---
output: github_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/castanedaM/nicheR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/castanedaM/nicheR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nicheR
### Ellipsoid-based virtual niches in E-space and G-space

**nicheR** is an R package for building, visualizing, predicting, and
sampling **ellipsoid-based ecological niches** using environmental data.

It provides:

- Explicit ellipsoid construction in n-dimensional environmental space
- Efficient suitability detection using Mahalanobis distance
- Flexible virtual occurrence sampling strategies (centroid, edge, random)
- Bias surface preparation and application for realistic sampling simulation
- Visualization tools in environmental space (E-space) and geographic space (G-space)
- A unified, fully scriptable R-native workflow

Inspired by the conceptual foundations of **NicheA** and the flexibility
of the **virtualspecies** package, **nicheR** provides a reproducible
framework for virtual species simulation and niche theory exploration.

---

## Authors

- Mariana Castaneda-Guzman
- Connor Hughes
- Paanwaris Paansri
- Marlon Cobos

---

## Installation

Install the development version from GitHub:

```{r message=FALSE, warning=FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("castanedaM/nicheR")
library(nicheR)
```

---

## Why nicheR?

The field of distributional ecology is evolving rapidly, with new algorithms,
parameterization strategies, and hypotheses emerging at a fast pace. Evaluating
these ideas requires software that can isolate mechanisms, account for bias, and
clearly link ecological theory to model behavior.

Virtual niche simulations help researchers test ideas by creating controlled
datasets that make model behavior easier to interpret. In practice, building
ellipsoid-based virtual niches in R has typically required combining several
packages to define niches, map them to geography, and visualize results —
making analyses harder to reproduce and build on.

**nicheR** addresses this gap by offering a single, integrated, R-native
framework that streamlines virtual niche construction, prediction, visualization,
and occurrence data generation into one reproducible workflow.

### Related work

Three tools have shaped virtual species and niche simulation in ENM/SDMs:

- **NicheA** — a Java-based software that pioneered environmental-space niche
  visualization and explicitly linked niche theory with simulation. An important
  conceptual foundation for nicheR.
- **[virtualspecies](https://github.com/Farewe/virtualspecies)** — a widely adopted R package (200+ citations) for simulating
  virtual species and benchmarking SDMs and ENMs within R.
- **[evniche](https://github.com/marlonecobos/evniche/)** — a theoretically rigorous R package focused on environmental-space
  representations and ellipsoid-based niche concepts.

nicheR builds directly on these approaches, unifying their strengths in a
well-documented R package. A Shiny GUI for interactive niche exploration is
also under active development.

---

## Conceptual Framework

nicheR operates across two complementary spaces:

- **E-space (Environmental Space)**: Ellipsoids represent multivariate
  environmental tolerances. This is where niches are defined, visualized,
  and compared.

- **G-space (Geographic Space)**: Predictions are projected across raster
  layers to map suitable geographic regions and generate virtual occurrence
  data.

This dual-space structure allows explicit separation between niche
definition, projection, and sampling processes, making it straightforward
to test ecological hypotheses under controlled, reproducible conditions.

---

## Workflow Overview

A typical nicheR workflow follows these steps:

1. **Prepare environmental data** — load and subset raster layers
2. **Prepare bias layers** *(optional)* — define sampling biases with directional effects
3. **Build an ellipsoid** — define the niche from environmental ranges
4. **Predict suitability** — project the ellipsoid onto raster or data frame inputs
5. **Apply bias** *(optional)* — weight predictions by a composite bias surface
6. **Sample occurrences** — draw virtual occurrence records using various strategies

---

## Quick Start

```{r}
library(nicheR)
library(terra)

# Load environmental layers, here we use three WorldClim bioclimatic variables
# for the Mesoamerica region included as example data in the package
bios <- terra::rast(system.file("extdata/ma_bios.tif", package = "nicheR"))
bios <- bios[[c("bio_1", "bio_12", "bio_15")]]

# Convert to data frame for E-space operations (keeping x/y coordinates)
bios_df <- as.data.frame(bios, xy = TRUE)
```

---

## Preparing Bias Layers

Bias layers represent external factors that influence sampling detectability
or effort, independent of the species' actual niche. Each layer is assigned
a direction: `"direct"` means higher values increase sampling probability
(e.g., higher species richness → most likely good habitat), while `"inverse"` means
higher values decrease it (e.g., higher light pollution → more urbanization → less likely to have good habitat).

`prepare_bias()` standardizes all layers to [0, 1], applies the directional
transformations, and combines them into a single composite bias surface.
Setting `include_processed_layers = TRUE` returns the standardized
individual layers alongside the composite, which is useful for inspecting
how each layer contributed.

```{r}
bias <- prepare_bias(bias_surface             = terra::rast(system.file("extdata/ma_biases.tif", package = "nicheR")),
                     effect_direction         = c("direct", "inverse"),
                     include_processed_layers = TRUE)

# Inspect the formula used to combine layers and the resulting composite
bias$combination_formula
bias$composite_surface

# Visualize: standardized individual layers and the composite bias surface
# The composite reflects both the direct (species richness) and inverse
# (nighttime light) effects combined into a single sampling probability surface
terra::plot(bias$processed_layers)
terra::plot(bias$composite_surface)
```

> **Vignette:** For a detailed walkthrough of bias surface construction,
> directional effects, and how bias interacts with predictions and sampling,
> see: *[insert link later Bias: preparation, prediction, and sampling]*

---

## Building an Ellipsoid

The ellipsoid is the core object in nicheR. It represents the virtual
species' niche as a multivariate ellipse in environmental space, defined
by a centroid and a covariance matrix that determine its position, size,
and orientation. At minimum, you define it with a range data frame, two
rows giving the minimum and maximum for each environmental variable.
nicheR computes the centroid, covariance matrix, semi-axes lengths, and
niche volume automatically.

```{r}
range <- data.frame(bio_1  = c(27, 35),
                    bio_12 = c(1000, 1500),
                    bio_15 = c(60, 75))

ell <- build_ellipsoid(range = range)

# The ellipsoid object contains niche geometry and metrics
names(ell)

# Pretty-printed summary of the ellipsoid
ell
```

### Visualizing the Ellipsoid

nicheR includes base-R plotting functions for visualizing the ellipsoid
in E-space. You can plot individual dimension pairs or all pairs at once.
Adding the environmental background helps contextualize where the niche
sits relative to available conditions, in this example the ellipsoid
occupies a warm (bio_1: 20–25), moderately seasonal corner of the broader
Mesoamerican climate space.

```{r}
# Basic E-space plots, no background, shows ellipsoid geometry only
plot_ellipsoid(ell)
plot_ellipsoid_pairs(ell)

# With environmental background, bg_sample controls how many background
# points are drawn for speed; increase for publication-quality plots
plot_ellipsoid(ell, background = bios_df, dim = c(1, 2), bg_sample = 5000)
plot_ellipsoid(ell, background = bios_df, dim = c(1, 3), bg_sample = 5000)

# Pairwise plot showing all dimension combinations simultaneously
plot_ellipsoid_pairs(ell,
                     background = bios_df, 
                     bg_sample = 5000,
                     col_ell = "red")
```

> **Vignette:** For details on ellipsoid construction, covariance-based
> rotation, niche volume, and advanced visualization options, see:
> *[insert link later, Building ellipsoids]*

---

## Predicting Suitability

Once the ellipsoid is built, `predict()` projects it onto environmental
data to produce suitability and Mahalanobis distance surfaces. Suitability
is highest at the niche centroid and decreases toward the boundary.
Truncated outputs (`_trunc`) set values to zero (for suitbaility) or NA (for Mahalanobis) outside the ellipsoid,
which is useful for enforcing strict niche-boundary sampling downstream.

Both raster (`SpatRaster`) and data frame inputs are supported and return
matching output types.

### Raster input

```{r}
# Returns a SpatRaster with: bio layers (from keep_data = TRUE),
# Mahalanobis distance, suitability, and their truncated versions
ell_predict_r <- predict(ell,
                         newdata               = bios,
                         suitability_truncated = TRUE,
                         mahalanobis_truncated = TRUE,
                         keep_data             = TRUE)

terra::plot(ell_predict_r)
```

### Data frame input

```{r}
# Returns the original data frame with appended prediction columns
ell_predict_df <- predict(ell,
                          newdata               = bios_df,
                          suitability_truncated = TRUE)

head(ell_predict_df)
```

### Visualizing predictions in E-space

Plotting predictions in E-space makes it easy to see how suitability
is distributed relative to the niche boundary. In the non-truncated view,
all background points are colored by suitability, showing a gradient
from low to high as points approach the ellipsoid center. The truncated
view zooms into the ellipsoid itself, revealing only the environmentally
suitable area with the suitability gradient inside.

```{r}
# Non-truncated: full background colored by suitability gradient
plot_ellipsoid(ell, background = ell_predict_df, pch = 20, bg_sample = 5000)
add_data(data      = ell_predict_df, x = "bio_1", y = "bio_12",
         col_layer = "suitability",
         rev_pal   = TRUE,
         pal       = hcl.colors(100, palette = "Viridis"),
         bg_sample = 10000,
         pch       = 20)
add_ellipsoid(ell, col_ell = "red")

# Truncated: only points inside the ellipsoid are shown, colored by suitability
plot_ellipsoid(ell)
add_data(data      = ell_predict_df, x = "bio_1", y = "bio_12",
         col_layer = "suitability_trunc",
         rev_pal   = TRUE,
         pch       = 20)
add_ellipsoid(ell, col_ell = "red", lwd = 2)


plot_ellipsoid_pairs(ell, prediction = ell_predict_df,
                     bg_sample = 5000, pch = 20, 
                     col_layer = "suitability")




```

> **Vignette:** For a full guide to prediction outputs, truncation logic,
> raster vs. data frame workflows, and E-space vs. G-space visualization,
> see: *[insert link later, Prediction and sampling]*

---

## Applying Bias to Predictions

`apply_bias()` weights the suitability prediction by the composite bias
surface to produce a biased suitability layer. This simulates how real
occurrence records are collected under imperfect, non-random sampling —
where detection probability is influenced by external factors like
accessibility, observer effort, or habitat visibility. Note that the
resulting layer is no longer a strict probability surface, but reflects
the joint effect of niche suitability and sampling bias.

```{r}
biased_predict_r <- apply_bias(prepared_bias    = bias,
                               prediction       = ell_predict_r,
                               prediction_layer = "suitability")
```

> **Vignette:** For details on bias-weighted prediction, comparing biased
> vs. unbiased outputs, and downstream effects on occurrence sampling, see:
> *[insert link later, Bias: preparation, prediction, and sampling]*

---

## Sampling Virtual Occurrences

nicheR provides two sampling functions: `sample_data()` for unbiased
sampling directly from suitability or Mahalanobis surfaces, and
`sample_biased_data()` for sampling from a bias-weighted prediction surface.

Three **sampling strategies** control where within the niche occurrences
are drawn from:

- `"centroid"` — samples preferentially near the niche center (most suitable conditions)
- `"edge"` — samples preferentially near the niche boundary (marginal conditions)
- `"random"` — samples uniformly across the suitable area

Two **weighting methods** control the probability used to draw samples:

- `"suitability"` — weights by suitability score
- `"mahalanobis"` — weights by Mahalanobis distance from the centroid

Using a truncated prediction layer with `strict = TRUE` ensures all
sampled points fall strictly within the ellipsoid boundary.

### Unbiased sampling — centroid strategy

```{r}
sample_data_df_cn <- sample_data(n_occ            = 100,
                                 prediction       = ell_predict_df,
                                 prediction_layer = "suitability_trunc",
                                 sampling         = "centroid",
                                 method           = "mahalanobis",
                                 strict           = TRUE)
head(sample_data_df_cn)
```

### Unbiased sampling — edge strategy

```{r}
sample_data_df_ed <- sample_data(n_occ            = 100,
                                 prediction       = ell_predict_df,
                                 prediction_layer = "suitability_trunc",
                                 sampling         = "edge",
                                 method           = "suitability")
```

### Biased sampling

```{r}
# Sampling from the bias-weighted suitability surface
# Even with raster input, the output is always a data frame of occurrence records
sample_data_b <- sample_biased_data(n_occ            = 100,
                                    prediction       = biased_predict_r,
                                    prediction_layer = "suitability_biased_direct")
head(sample_data_b)
```

### Comparing centroid and edge sampling

The centroid strategy concentrates records near the niche center, while
the edge strategy spreads records toward the ellipsoid boundary. Both are
plotted against the environmental background to illustrate the difference
in spatial distribution within E-space, black crosses (×) are the sampled
occurrences, grey points are the background, and the red ellipse is the
niche boundary.

```{r}
par(mfrow = c(2, 1))

# Centroid sampling, occurrences cluster toward the niche center
plot_ellipsoid(ell, main = "Centroid Sampling — Unbiased")
add_data(data    = ell_predict_df, x = "bio_1", y = "bio_12",
         pts_col = "grey", pch = 20, bg_sample = 8000)
add_data(data    = sample_data_df_cn, x = "bio_1", y = "bio_12",
         pts_col = "black", pch = 4, lwd = 2)
add_ellipsoid(ell, col_ell = "red", lwd = 2)

# Edge sampling, occurrences spread toward the niche boundary
plot_ellipsoid(ell, main = "Edge Sampling — Unbiased")
add_data(data    = ell_predict_df, x = "bio_1", y = "bio_12",
         pts_col = "grey", pch = 20, bg_sample = 8000)
add_data(data    = sample_data_df_ed, x = "bio_1", y = "bio_12",
         pts_col = "black", pch = 4, lwd = 2)
add_ellipsoid(ell, col_ell = "red", lwd = 2)
```

> **Vignette:** For a full guide to sampling strategies, methods, strict
> vs. non-strict modes, and biased sampling workflows, see:
> *[insert link later, Prediction and sampling]*

---

## Hypothesis Testing

nicheR includes functions for testing ecological hypotheses using virtual
niche data. The community function allows users to evaluate hypotheses
about niche overlap, divergence, and community-level patterns by comparing
ellipsoids across species or conditions under controlled virtual scenarios.

> **Vignette:** For details on hypothesis testing with virtual niches and
> community-level comparisons, see:
> *[insert link later, Hypothesis testing: community niche analyses]*

---

## Coming Soon

Detailed vignettes are currently in development:

- *Building ellipsoids* — construction, covariance, rotation, and niche volume
- *Prediction and sampling* — raster and data frame workflows, truncation, sampling strategies
- *Bias* — preparation, directional effects, bias-weighted prediction and sampling
- *Hypothesis testing* — community niche analyses and virtual niche comparisons

An interactive **Shiny GUI** for point-and-click niche construction,
prediction, and occurrence sampling is also planned, with a focus on
preserving full reproducibility by exporting session-generated R scripts.

---

## Contributing

We welcome contributions. If you have suggestions, bug reports, feature
requests, or code improvements, please open an issue or submit a pull
request on GitHub.

When contributing:

- Document functions using roxygen2
- Include minimal reproducible examples
- Add tests when appropriate

Thank you for helping improve **nicheR**.
