---
title: "Sampling Occurrence Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling Occurrence Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  dpi = 100,
  out.width = "95%"
)
```

## Summary

- [Description](#description)
- [Getting ready](#getting-ready)
- [1. Defining the Niche Space: build_ellipsoid()](#1-defining-the-niche-space-build_ellipsoid)
- [2. Generating the Prediction Surface](#2-generating-the-prediction-surface)
- [3. Standard Occurrence Sampling: sample_data()](#3-standard-occurrence-sampling-sample_data)
  - [Suitability | centroid | Strict= FALSE](#suitability--centroid--strict--false)
  - [Mahalanobis | edge | Strict= FALSE](#mahalanobis--edge--strict--false)
  - [Suitability Truncated | random | Strict= TRUE](#suitability-truncated--random--strict--true)
  - [Mahalanobis Truncated | edge | Strict= TRUE](#mahalanobis-truncated--edge--strict--true)
- [4. Introducing Sampling Bias: sample_biased_data()](#4-introducing-sampling-bias-sample_biased_data)
  - [Prerequisites: Prepare the Bias Layer](#prerequisites-prepare-the-bias-layer)
  - [Suitability | Direct Bias | Strict= FALSE](#suitability--direct-bias--strict--false)
  - [Suitability Truncated | Inverse Bias | Strict= TRUE](#suitability-truncated--inverse-bias--strict--true)
  - [Mahalanobis | Direct Bias | Strict= FALSE](#mahalanobis--direct-bias--strict--false)
  - [Mahalanobis Truncated | Inverse Bias | Strict= TRUE](#mahalanobis-truncated--inverse-bias--strict--true)
- [5. Virtual Species Sampling: sample_virtual_data()](#5-virtual-species-sampling-sample_virtual_data)
  - [Suitability | Centroid | Strict= FALSE](#suitability--centroid--strict--false-1)
  - [Suitability Truncated | Random | Strict= TRUE](#suitability-truncated--random--strict--true-1)
  - [Mahalanobis | Edge | Strict= FALSE](#mahalanobis--edge--strict--false-1)
  - [Mahalanobis Truncated | Centroid | Strict= TRUE](#mahalanobis-truncated--centroid--strict--true)

<hr>

# Description

The **nicheR** package provides a flexible suite of functions for drawing
virtual occurrence points from suitability or Mahalanobis distance surfaces.
These functions cover three main workflows:

- **Standard spatial sampling** (`sample_data()`): samples from a raster or
  data frame prediction surface, returning points with geographic coordinates.
- **Biased spatial sampling** (`sample_biased_data()`): samples from a
  bias-weighted prediction surface to simulate non-random collection effort.
- **Virtual (non-spatial) sampling** (`sample_virtual_data()`): samples
  directly from a purely virtual prediction — no raster or geographic
  coordinates involved, working entirely in environmental space.

All three functions support the same sampling strategies (`"centroid"`,
`"edge"`, `"random"`) and weighting methods (`"suitability"`,
`"mahalanobis"`), and share a consistent interface for the `strict` argument
that controls how zero and `NA` values are handled.

<br>

# Getting ready

```{r}
library(nicheR)
library(terra)
```

<br>

# 1. Defining the Niche Space: `build_ellipsoid()`

We start by defining an ellipsoid niche using known environmental ranges
for three bioclimatic variables. The ellipsoid represents the set of
environmental conditions the virtual species can tolerate. All sampling
functions in this vignette are built on top of this object.

```{r}
bios_file <- system.file("extdata", "ma_bios.tif", package = "nicheR")
bios      <- rast(bios_file)
vars      <- c("bio_1", "bio_12", "bio_15")

# Define the environmental limits of the virtual niche
range_df <- data.frame(bio_1 = c(22, 28),
                       bio_12 = c(1000, 3500),
                       bio_15 = c(50, 70))

ell <- build_ellipsoid(range = range_df)
```

<br>

# 2. Generating the Prediction Surface

Before sampling, we project the ellipsoid onto the environmental raster to
get suitability and Mahalanobis distance values for every cell in the study
area. We generate both standard and truncated versions of each — truncated
layers set values to zero outside the ellipsoid boundary, which is useful
when you want to restrict sampling strictly to suitable conditions.

```{r}
pred <- predict(ell,
                newdata = bios,
                include_mahalanobis = TRUE,
                include_suitability = TRUE,
                suitability_truncated = TRUE,
                mahalanobis_truncated = TRUE)
plot(pred)
```

<br>

# 3. Standard Occurrence Sampling: `sample_data()`

`sample_data()` draws occurrence points from a suitability or Mahalanobis
distance prediction surface using weighted random sampling. This is the
main function for generating presence records under controlled, unbiased
conditions.

The `sampling` argument controls *where* within the niche points are
preferentially drawn from:

- `"centroid"` — concentrates points near the most suitable conditions
  (niche center).
- `"edge"` — spreads points toward the niche boundary (marginal conditions).
- `"random"` — samples uniformly, ignoring suitability gradients.

The `method` argument controls *how* prediction values are used as weights:

- `"suitability"` — uses values directly (must be in [0, 1]).
- `"mahalanobis"` — uses Mahalanobis distance (must be non-negative).

When using a truncated layer, set `strict = TRUE` to remove zeros and `NA`s
before sampling — this ensures all sampled points fall inside the ellipsoid.

<br>

## Suitability | centroid | Strict= FALSE

Points are drawn from the full suitability surface with higher probability
near the niche center. Since `strict = FALSE`, points outside the ellipsoid
boundary (low but non-zero suitability) are included in the candidate pool.

```{r}
occ_suit_cent <- sample_data(n_occ = 100,
                             prediction = pred,
                             prediction_layer = "suitability",
                             sampling = "centroid",
                             method = "suitability",
                             seed = 123,
                             strict = FALSE)

occ_suit_cent_env <- terra::extract(bios, occ_suit_cent[, c(1:2)]) |>
  cbind(occ_suit_cent[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Suitability | Centroid (Bio1 v Bio12)")
add_data(occ_suit_cent_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Suitability | Centroid (Bio12 v Bio15)")
add_data(occ_suit_cent_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis | edge | Strict= FALSE

Points are drawn using Mahalanobis distance with edge weighting — higher
probability is given to cells farther from the niche centroid. This simulates
occurrences recorded at the margins of the species' environmental tolerance.

```{r}
occ_maha_edge <- sample_data(n_occ = 100,
                             prediction = pred,
                             prediction_layer = "Mahalanobis",
                             sampling = "edge",
                             method = "mahalanobis",
                             seed = 123,
                             strict = FALSE)

occ_maha_edge_env <- terra::extract(bios, occ_maha_edge[, c(1:2)]) |>
  cbind(occ_maha_edge[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Mahalanobis | Edge (Bio1 v Bio12)")
add_data(occ_maha_edge_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Mahalanobis | Edge (Bio12 v Bio15)")
add_data(occ_maha_edge_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Suitability Truncated | random | Strict= TRUE

Using a truncated suitability layer with `strict = TRUE` ensures all sampled
points fall strictly inside the ellipsoid. The `"random"` strategy then
samples uniformly across that restricted suitable area — no preference for
center or edge.

```{r}
occ_suit_trunc_rand <- sample_data(n_occ = 100,
                                   prediction = pred,
                                   prediction_layer = "suitability_trunc",
                                   sampling = "random",
                                   method = "suitability",
                                   seed = 123,
                                   strict = TRUE)

occ_suit_trunc_rand_env <- terra::extract(bios, occ_suit_trunc_rand[, c(1:2)]) |>
  cbind(occ_suit_trunc_rand[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Suit Truncated | Random (Bio1 v Bio12)")
add_data(occ_suit_trunc_rand_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Suit Truncated | Random (Bio12 v Bio15)")
add_data(occ_suit_trunc_rand_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis Truncated | edge | Strict= TRUE

Combining a truncated Mahalanobis layer with edge sampling and `strict = TRUE`
produces points that are both inside the ellipsoid and biased toward its
outer boundary — useful for simulating species detected at the limits of
their tolerance range.

```{r}
occ_maha_trunc_edge <- sample_data(n_occ = 100,
                                   prediction = pred,
                                   prediction_layer = "Mahalanobis_trunc",
                                   sampling = "edge",
                                   method = "mahalanobis",
                                   seed = 123,
                                   strict = TRUE)

occ_maha_trunc_edge_env <- terra::extract(bios, occ_maha_trunc_edge[, c(1:2)]) |>
  cbind(occ_maha_trunc_edge[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Maha Truncated | Edge (Bio1 v Bio12)")
add_data(occ_maha_trunc_edge_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Maha Truncated | Edge (Bio12 v Bio15)")
add_data(occ_maha_trunc_edge_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)

dev.off()
```

<br>

# 4. Introducing Sampling Bias: `sample_biased_data()`

Real-world biodiversity records are rarely collected evenly across space —
surveys cluster near roads, research stations, and accessible areas, and
detectability varies with habitat type or observer effort. `sample_biased_data()`
simulates this by sampling from a prediction layer that has been weighted
by an external bias surface.

The bias surface is prepared with `prepare_bias()`, which standardizes each
bias layer to [0, 1] and assigns a directional effect:

- `"direct"` — higher values increase sampling probability (e.g., higher
  species richness → more survey effort).
- `"inverse"` — higher values decrease sampling probability (e.g., higher
  nighttime light → fewer detections).

The resulting composite bias surface is then combined with the suitability
or Mahalanobis prediction using `apply_bias()` before sampling.

## Prerequisites: Prepare the Bias Layer

```{r}
biases_file <- system.file("extdata", "ma_biases.tif", package = "nicheR")
biases      <- rast(biases_file)

prep_biases <- prepare_bias(bias_surface = biases,
                            effect_direction = c("direct", "inverse"))
```

<br>

## Suitability | Direct Bias | Strict= FALSE

The suitability surface is weighted by the direct bias — cells with higher
bias values are more likely to be sampled. This shifts occurrence points
toward areas of high sampling effort regardless of suitability alone.

```{r}
app_suit_dir <- apply_bias(prepared_bias = prep_biases,
                           prediction = pred,
                           prediction_layer = "suitability",
                           effect_direction = "direct")

occ_suit_dir <- sample_biased_data(n_occ = 100,
                                   prediction = app_suit_dir,
                                   prediction_layer = "suitability_biased_direct",
                                   seed = 123,
                                   strict = FALSE)

occ_suit_dir_env <- terra::extract(bios, occ_suit_dir[, c(1:2)]) |>
  cbind(occ_suit_dir[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Suitability | Direct Bias (1v12)")
add_data(occ_suit_dir_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Suitability | Direct Bias (12v15)")
add_data(occ_suit_dir_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Suitability Truncated | Inverse Bias | Strict= TRUE

Using the truncated suitability layer restricts sampling to inside the
ellipsoid, while the inverse bias further shifts points away from high-bias
areas — simulating a scenario where high-effort areas are paradoxically
under-sampled (e.g., urban light pollution suppressing nocturnal detections).

```{r}
app_suit_trunc_inv <- apply_bias(prepared_bias = prep_biases,
                                 prediction = pred,
                                 prediction_layer = "suitability_trunc",
                                 effect_direction = "inverse")

occ_suit_trunc_inv <- sample_biased_data(n_occ = 100,
                                         prediction = app_suit_trunc_inv,
                                         prediction_layer = "suitability_trunc_biased_inverse",
                                         strict = TRUE)

occ_suit_trunc_inv_env <- terra::extract(bios, occ_suit_trunc_inv[, c(1:2)]) |>
  cbind(occ_suit_trunc_inv[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Suit Truncated | Inverse Bias (1v12)")
add_data(occ_suit_trunc_inv_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Suit Truncated | Inverse Bias (12v15)")
add_data(occ_suit_trunc_inv_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis | Direct Bias | Strict= FALSE

Mahalanobis distance is used as the base prediction, weighted by the direct
bias surface. Points are drawn from cells that are both far from the centroid
(high Mahalanobis distance) and in high-bias areas — edge-of-niche conditions
under biased detection.

```{r}
app_maha_dir <- apply_bias(prepared_bias = prep_biases,
                           prediction = pred,
                           prediction_layer = "Mahalanobis",
                           effect_direction = "direct")

occ_maha_dir <- sample_biased_data(n_occ = 100,
                                   prediction = app_maha_dir,
                                   prediction_layer = "Mahalanobis_biased_direct",
                                   strict = FALSE)

occ_maha_dir_env <- terra::extract(bios, occ_maha_dir[, c(1:2)]) |>
  cbind(occ_maha_dir[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Mahalanobis | Direct Bias (1v12)")
add_data(occ_maha_dir_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Mahalanobis | Direct Bias (12v15)")
add_data(occ_maha_dir_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis Truncated | Inverse Bias | Strict= TRUE

All three restrictions combined: truncation keeps points inside the ellipsoid,
the inverse bias downweights high-effort areas, and `strict = TRUE` removes
any residual zeros or `NA`s. This is the most conservative scenario —
sampling is restricted to suitable conditions but avoids areas where detection
is most likely.

```{r}
app_maha_trunc_inv <- apply_bias(prepared_bias = prep_biases,
                                 prediction = pred,
                                 prediction_layer = "Mahalanobis_trunc",
                                 effect_direction = "inverse")

occ_maha_trunc_inv <- sample_biased_data(n_occ = 100,
                                         prediction = app_maha_trunc_inv,
                                         prediction_layer = "Mahalanobis_trunc_biased_inverse",
                                         strict = TRUE)

occ_maha_trunc_inv_env <- terra::extract(bios, occ_maha_trunc_inv[, c(1:2)]) |>
  cbind(occ_maha_trunc_inv[3])

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Maha Truncated | Inverse Bias (1v12)")
add_data(occ_maha_trunc_inv_env,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(ell,
               background = as.data.frame(bios[[vars]]),
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Maha Truncated | Inverse Bias (12v15)")
add_data(occ_maha_trunc_inv_env,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)

dev.off()
```

<br>

# 5. Virtual Species Sampling: `sample_virtual_data()`

`sample_virtual_data()` is designed for purely virtual workflows — it samples
from a prediction built entirely in environmental (non-geographic) space.
Instead of projecting onto a raster, we generate a large pool of virtual
environmental points using `virtual_data()`, predict over those points, and
then sample from the resulting data frame.

Because there are no geographic coordinates, there is no `terra::extract()`
step — sampled points already contain the environmental values directly, and
can be plotted straight from the sampling output.

```{r}
virtual_range <- data.frame(bio_1 = c(20, 29),
                            bio_12 = c(500, 4000),
                            bio_15 = c(30, 100))

virtual_ell  <- build_ellipsoid(range = virtual_range)

# Generate a pool of virtual environmental points from the ellipsoid,
# then predict suitability over them — no raster involved
virtual_pred <- predict(virtual_ell,
                        newdata = virtual_data(virtual_ell, n = 10000),
                        include_mahalanobis = TRUE,
                        include_suitability = TRUE,
                        suitability_truncated = TRUE,
                        mahalanobis_truncated = TRUE)
```

<br>

## Suitability | Centroid | Strict= FALSE

Points are drawn from the virtual suitability surface with centroid weighting.
The output is a data frame with environmental values — no extraction needed.

```{r}
occ_virt_suit_cent <- sample_virtual_data(n_occ = 100,
                                          object = virtual_ell,
                                          virtual_prediction = virtual_pred,
                                          prediction_layer = "suitability",
                                          sampling = "centroid",
                                          method = "suitability",
                                          seed = 123,
                                          strict = FALSE)

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(virtual_ell,
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Virtual Suitability | Centroid (1v12)")
add_data(occ_virt_suit_cent,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(virtual_ell,
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Virtual Suitability | Centroid (12v15)")
add_data(occ_virt_suit_cent,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Suitability Truncated | Random | Strict= TRUE

Random sampling from the truncated suitability surface — all points are
inside the ellipsoid and drawn with equal probability.

```{r}
occ_virt_suit_trunc_rand <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "suitability_trunc",
  sampling = "random",
  method = "suitability",
  seed = 123,
  strict = TRUE
)

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(virtual_ell,
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Virtual Suit Trunc | Random (1v12)")
add_data(occ_virt_suit_trunc_rand,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(virtual_ell,
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Virtual Suit Trunc | Random (12v15)")
add_data(occ_virt_suit_trunc_rand,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis | Edge | Strict= FALSE

Edge weighting using Mahalanobis distance pushes points toward the boundary
of the virtual niche. Since `strict = FALSE`, points slightly outside the
ellipsoid are also eligible.

```{r}
occ_virt_maha_edge <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "Mahalanobis",
  sampling = "edge",
  method = "mahalanobis",
  seed = 123,
  strict = FALSE
)

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(virtual_ell,
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Virtual Mahalanobis | Edge (1v12)")
add_data(occ_virt_maha_edge,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(virtual_ell,
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Virtual Mahalanobis | Edge (12v15)")
add_data(occ_virt_maha_edge,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)
```

<br>

## Mahalanobis Truncated | Centroid | Strict= TRUE

Centroid weighting with a truncated Mahalanobis layer and `strict = TRUE` —
all points are inside the ellipsoid and concentrated near the niche center.
This is the most controlled scenario for virtual species simulation.

```{r}
occ_virt_maha_trunc_cent <- sample_virtual_data(
  n_occ = 100,
  object = virtual_ell,
  virtual_prediction = virtual_pred,
  prediction_layer = "Mahalanobis_trunc",
  sampling = "centroid",
  method = "mahalanobis",
  seed = 123,
  strict = TRUE
)

# --- Plotting ---
par(mfrow = c(1, 2))

plot_ellipsoid(virtual_ell,
               dim = c(1, 2),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio1",
               ylab = "Bio12",
               main = "Virt Maha Trunc | Centroid (1v12)")
add_data(occ_virt_maha_trunc_cent,
         x = "bio_1",
         y = "bio_12",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_1",
         y = "bio_12",
         pts_col = "red",
         pch = 15)

plot_ellipsoid(virtual_ell,
               dim = c(2, 3),
               pch = ".",
               col_bg = "#9a9797",
               xlab = "Bio12",
               ylab = "Bio15",
               main = "Virt Maha Trunc | Centroid (12v15)")
add_data(occ_virt_maha_trunc_cent,
         x = "bio_12",
         y = "bio_15",
         pts_col = "orange",
         pch = 20)
add_data(as.data.frame(t(virtual_ell$centroid)),
         x = "bio_12",
         y = "bio_15",
         pts_col = "red",
         pch = 15)

dev.off()
```
