---
title: "Virtual community simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Virtual community simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

## Summary

- [Description](#description)
- [Getting ready](#getting-ready)
- [Loading example data](#loading-example-data)
- [Simulating a random community](#simulating-a-random-community)
  - [Effect of background density](#effect-of-background-density)
  - [Effect of proportion arguments](#effect-of-proportion-arguments)
- [Simulating nested communities](#simulating-nested-communities)
  - [Effect of proportion argument](#effect-of-proportion-argument)
  - [Effect of bias argument](#effect-of-bias-argument)
- [Niche conservatism in communities](#niche-conservatism-in-communities)
  - [Effect of background density](#effect-of-background-density)
  - [Effect of proportion arguments](#effect-of-proportion-arguments)
- [Save and import communities](#save-and-import-communities)

<hr>

## Description

This vignette shows how to simulate virtual communities using functions in the nicheR package. For our purposes, we will consider a community a set of species niches (ellipses) that are distributed in a given environmental space. The simulation of virtual communities is useful to generate hypothetical scenarios of community assembly and explore patterns derived from the way species are distributed in envrionmental and geographical space. 

The main functions that automate community simulations in `nicheR` are:

- `random_ellipses()`: generates a set of random ellipses in the environmental space provided. A scenario in which niche conservatism is not present.
- `nested_ellipses()`: generates a set of nested ellipses taking the reference niche as the biggest possible ellipse. A scenario in which all species niches are contained within the reference niche, and each is contained within the previous one.
- `conserved_ellipses()`: generates a set of ellipses in the environmental space provided aiming for a set of ellipses that are similar to the reference niche. A scenario of niche conservatism.

<br>

## Getting ready

If `nicheR`  has not been installed yet, please do so. See the [Main guide](index.html) for installation instructions.

Use the following lines of code to load `nicheR` and other packages needed for this vignette, and to set a working directory (if necessary).

Note: We will display functions from other packages as `package::function()`. 

```{r get_ready, results='hide', message=FALSE, warning=FALSE}
# Load packages
library(nicheR)
library(terra)

# Current directory
getwd()

# Define new directory
#setwd("YOUR/DIRECTORY")  # uncomment and modify if setting a new directory

# Saving original plotting parameters
original_par <- par(no.readonly = TRUE)
```

<br>

## Loading example data

The lines of code below help us load the data to run our example comunity simulations. The data are included in the `nicheR` package and consist of an `nicheR_ellipsoid` object with a reference niche defined by two environmental variables (bio1 and bio12) and these envrionmental variables for North America to represent the background for our simulations.

```{r data}
# Reference niche
data("ref_ellipse", package = "nicheR")

# Background data
data("back_data", package = "nicheR")
```

<br>

Let's take a look at the reference niche and the background data to get familiar with them before we start simulating communities.
```{r data_check}
# Check reference niche
print(ref_ellipse)

# Check background data
head(back_data)

# Plotting the background data to visualize the environmental space
par(mar = c(4, 4, 2, 1))  # adjust margins for better visualization

plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Background Environmental Space and Reference Niche")

# Adding the reference niche to the plot
add_ellipsoid(ref_ellipse, col = "#000000", lwd = 2)
```

<br>


## Simulating random communities

The `random_ellipses()` function helps to create a community of species with niches that are randomly distributed in the environmental space. The size and shape of these niches are constrained by the reference niche but they vary. Below we show an general example of how to use this function.

```{r random}
# Simulating the community
rand_comm <- random_ellipses(object = ref_ellipse,
                             background = back_data[, c("bio_1", "bio_12")],
                             n = 20)  # number of species in the community

# check the a few details from the generated community
names(rand_comm)  # elements in the community object

print(rand_comm)  # a summary of the elements in the community object
```

<br> 

Now let's us plot the generated community to visualize the distribution of the random ellipses in environmental space. We will use the `add_ellipsoid()` function to add the ellipses to a plot of the background data.

```{r random_plot}
# Plotting the community
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Background and Community of Random Ellipses")

## defined colors for ellispes
cols <- rainbow(length(rand_comm$ellipse_community))

## Add random ellipses in loop
for (i in seq_along(rand_comm$ellipse_community)) {
  add_ellipsoid(rand_comm$ellipse_community[[i]], col = cols[i])
}
```

<br>

### Effect of background density

As the `random_ellipses()` function uses the background as a reference to pick ellipse centroids, the density of the background data can affect the distribution of these ellipses. If the background data is dense in certain areas of the environmental space, it may lead to a higher concentration of ellipses in those areas, while sparse background data may result in fewer ellipses being generated in those regions. This can influence the overall structure and diversity of the simulated community, as well as the patterns observed in niche overlap and species interactions.

The example below shows how the density of the background can affect the distribution of random ellipses in the environmental space. We will simulate two communities: (1) using the full background as reference, and (2) using the arguments thin_background and resolution to reduce the effect of uneven point density. We will increase the number of species in the community to better visualize the effect of background density on the distribution of ellipses.

```{r random_den}
# Simulating the community with the full background
rand_comm_full <- random_ellipses(object = ref_ellipse,
                                  background = back_data[, c("bio_1", "bio_12")],
                                  n = 25)

# Simulating the community with a thinned background
rand_comm_thin <- random_ellipses(object = ref_ellipse,
                                  background = back_data[, c("bio_1", "bio_12")],
                                  n = 25, thin_background = TRUE,
                                  resolution = 20)

# Check summary of the two communities
rand_comm_full

rand_comm_thin
```

<br>

Let's check the distribution of ellipses in the two communities with a plot to visualize the effect of the arguments `thin_background` and `resolution`.

```{r random_den_plot, fig.width = 7, fig.height = 3.5, out.width = "95%"}  
# Plotting the communities
## defined colors for ellipses
cols <- rainbow(length(rand_comm_full$ellipse_community))

par(mfrow = c(1, 2), cex = 0.6)  # set up the plotting area

## Plotting the community with the full background
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Random Community from Full Background")

## Add random ellipses in loop
for (i in seq_along(rand_comm_full$ellipse_community)) {
  add_ellipsoid(rand_comm_full$ellipse_community[[i]], col = cols[i])
}

## Plotting the community with the thinned background
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Random Community from Thinned Background")

## Add random ellipses in loop
for (i in seq_along(rand_comm_thin$ellipse_community)) {
  add_ellipsoid(rand_comm_thin$ellipse_community[[i]], col = cols[i])
}
```

<br>

### Effect of proportion arguments

The size of the ellipses in the communities is determined by the reference niche and the arguments `smallest_proportion` and `largest_proportion` in the `random_ellipses()` function. Based on the reference niche, playing with the values for the areguments `smallest_proportion` and `largest_proportion` can help explore scenarios with . This can affect elements of the community structure such as the level of niche overlap observed among the species (ellipses).

The example below shows how these proportions can affect the coverage and overlap of random ellipses in environmental space. Considering that `smallest_proportion` must be smaller than `largest_proportion`, we will simulate four communities: (1) both arguments are small; and (2) both arguments are large; (3) `smallest_proportion` is large and `largest_proportion` is small; and (4) `smallest_proportion` is small and `largest_proportion` is large. We will thin the background in both cases to reduce the effect of uneven point density and better visualize the effect of the argument `smaller_proportion`.

```{r random_prop}
# Community with both arguments small
rand_comm1 <- random_ellipses(object = ref_ellipse,
                              background = back_data[, c("bio_1", "bio_12")],
                              n = 25, thin_background = TRUE,
                              resolution = 20,
                              smallest_proportion = 0.1,
                              largest_proportion = 0.5)

# Community with both arguments large
rand_comm2 <- random_ellipses(object = ref_ellipse,
                              background = back_data[, c("bio_1", "bio_12")],
                              n = 25, thin_background = TRUE,
                              resolution = 20,
                              smallest_proportion = 0.7,
                              largest_proportion = 1.5)

# Community with smallest_proportion large and largest_proportion small
rand_comm3 <- random_ellipses(object = ref_ellipse,
                              background = back_data[, c("bio_1", "bio_12")],
                              n = 25, thin_background = TRUE,
                              resolution = 20,
                              smallest_proportion = 0.5,
                              largest_proportion = 0.7)

# Community with smallest_proportion small and largest_proportion large
rand_comm4 <- random_ellipses(object = ref_ellipse,
                              background = back_data[, c("bio_1", "bio_12")],
                              n = 25, thin_background = TRUE,
                              resolution = 20,
                              smallest_proportion = 0.1,
                              largest_proportion = 1.5)

# Check summary of the communities
rand_comm1

rand_comm2

rand_comm3

rand_comm4
```

<br>

Let's check the distribution of ellipses in the two communities with a plot to visualize the effect of the argument `smaller_proportion`.

```{r random_prop_plot}
# Plotting the communities
## defined colors for ellipses
cols <- rainbow(length(rand_comm1$ellipse_community))

par(mfrow = c(2, 2), cex = 0.6)  # set up the plotting area

## Community with small proportions
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Random Community with Small Proportions")

## Add random ellipses in loop
for (i in seq_along(rand_comm1$ellipse_community)) {
  add_ellipsoid(rand_comm1$ellipse_community[[i]], col = cols[i])
}

## Community with large proportions
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Random Community with Large Proportions")

## Add random ellipses in loop
for (i in seq_along(rand_comm2$ellipse_community)) {
  add_ellipsoid(rand_comm2$ellipse_community[[i]], col = cols[i])
}

## Community with large smallest_proportion and small largest_proportion
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Random Community with Large and Small")

## Add random ellipses in loop
for (i in seq_along(rand_comm3$ellipse_community)) {
  add_ellipsoid(rand_comm3$ellipse_community[[i]], col = cols[i])
}

## Community with small smallest_proportion and large largest_proportion
plot(back_data[, c("bio_1", "bio_12")], pch = ".", col = "#9a9797",
     xlab = "Bio1 (Mean Annual Temperature)",
     ylab = "Bio12 (Annual Precipitation)",
     main = "Random Community with Small and Large")

## Add random ellipses in loop
for (i in seq_along(rand_comm4$ellipse_community)) {
  add_ellipsoid(rand_comm4$ellipse_community[[i]], col = cols[i])
}
```

<br> 

## Simulating nested communities

The `nested_ellipses()` function helps to create a community of species with niches that are nested in the environmental space. The size and shape of these niches are constrained by the reference niche but they vary. Below we show an general example of how to use this function.

<br>

### Effect of proportion argument

<br>

### Effect of bias argument

<br>

## Niche conservatism in communities

<br>

### Effect of background density

<br>

### Effect of proportion arguments

<br>

<br> 

```{r par_reset}
# Reset plotting parameters
par(original_par)
```

<br>

## Save and import communities
